---
title: "File_Download_and_RDS_Data_Inspection.Rmd"
author: "Myrto"
date: "2026-02-23"
output: html_document
---

This document provides code that can be used to generate the figure panels of the manuscript

#Set the environment
```{r setEnvironment, purl=TRUE}
library(Seurat)
library(ggplot2)
library(tidyverse)
library(SeuratDisk)
```
# ###############################################################################
# Downloading the files

```{r}
dest_file <- "PATH_TO_OUTPUT_FOLDER" #Replace with the path of the folder in quotes where all of the data will be saved
file_name <- "FILE_NAME" #Replace with the name of the file in quotes.
```

# 1. Downloading Seurat Objects from GEO (Recommended)
```{r}
url <- "URL_FROM_GEO" #Replace with the GEO URL in quotes of the file you want to download 
options(timeout = 600) #it indicates the number of seconds that are used as a threshold for each file download. The larger the file the longer it will take to download. Increase accordingly depending on the file size.
download.file(url, destfile = file.path(dest_file, file_name), mode = "wb") #downloads the file
```

# 2. Downloading the Seurat Objects from GEO Manually (Recommended)

# 3. Downloading loom files from Scope (NOT Recommended)
```{r}
#We recommend that you access the 'rds' files from our GEO page. However, if you wish to access the `.loom` file you can follow the following steps:
  #1. Downloading the `.loom` file from Scope
      #The `.loom` file can be downloaded from Scope (floppy disk icon) but it does NOT include the metadata information.
      #For more information you can take a look at the Tutorials in Scope: https://www.youtube.com/watch?v=tHUNMo3Bsi0&t=2s
  #2. Downloading the metadata 
      #The metadata of each .rds are saved on our GEO page. You can use the code in `1. Downloading Seurat Objects from GEO (Recommended)` to download the files or download them manually.

#Importing the data        
loom_path <- "PATH_TO_LOOM"    #Replace with the path of the folder in quotes where the loom was saved       
metadata_path <- file.path(dest_file, file_name) #If the metadata object was NOT downloaded through `1. Downloading Seurat Objects from GEO` above, replace the paths with the location of the file in your system

import_loom <- Connect(loom_path, mode = "r") #gets the loom
meta.data <- as.data.frame(read_csv(metadata_path)) 

dgem <- get_dgem(import_loom) #isolates information from the loom
Seurat_object <- CreateSeuratObject(counts = dgem, meta.data = meta.data) #reconstructs the Seurat Object

#saving the object
saveRDS(Seurat_object, file = file.path(dest_file, file_name))

```
# ###############################################################################
# Importing the data and setting the parameters

```{r Figure1, purl=TRUE}

outputFolder <- "PATH_TO_OUTPUT_FOLDER" #Replace with the path of the folder where all of the figures will be saved

#Importing the .rds objects
#If the objects were NOT downloaded through `1. Downloading Seurat Objects from GEO` above, replace the paths with the location of the files in your system
VNC.all <- file.path(dest_file, file_name) 
VNC.neurons <- file.path(dest_file, file_name)
VNC.subset.1ary <- file.path(dest_file, file_name)
VNC.subset.2ary <- file.path(dest_file, file_name)
```

# ###############################################################################
# Analysis

## Glia annotation 
See UMAP in panel 1b for more information
```{r}
cols.glia.annotation <- c(wesanderson::wes_palettes[["Darjeeling1"]][1:3])
names(cols.glia.annotation) <- c("glia","neuron","other")

DimPlot(VNC.all, 
        reduction = "UMAPallages", 
        group.by = "CellType",
        cols = cols.glia.annotation, 
        pt.size = 1.5, 
        raster.dpi = c(1180,1180)) + 
labs(title = "Glia annotation")
ggsave(file.path(outputFolder, "glia_annotation.png"), width = 20, height = 17, units = "cm", dpi = 600)
```

## Table with the number of glia, neurons and other cell types in females and males
See table in panel 1b for more information
```{r}
#creating an empty matrix
matrix <- matrix(nrow=4, ncol=4, 0)
rownames(matrix) <- c("neuron", "glia", "other", "Total")
colnames(matrix) <- c("F", "F%", "M", "M%")

#adding the number of cells per cell type and sex
for (x in c("neuron", "glia", "other")){
  matrix[x,c("F", "M")] <-table(VNC.all$CellType, VNC.all$Sex)[x,]
}
#calculating total numbers and percentages
matrix["Total", c("F", "M")] <- c(sum(matrix[,"F"]), sum(matrix[,"M"]))
matrix [, "F%"] <- round(100*(matrix[,"F"]/matrix["Total", "F"]))
matrix [, "M%"] <- round(100*(matrix[,"M"]/matrix["Total", "M"]))
```

## UMAP of neurons BEFORE stage integration
See panel 1c for more information
```{r}

Stage.cols<- c("6"="#009E73", "24"="#CC79A7", "36"="#E69F00", "48"="#0072B2")
   
DimPlot(VNC.neurons, 
        cols = Stage.cols, 
        reduction = "UMAPallages", 
        group.by = "Stage", 
        pt.size = 1.5, 
        raster.dpi = c(1180,1180)) &
labs(title = "Neurons pre-Integration")
ggsave(file.path(outputFolder , "neurons_pre-integration.png"), width = 20, height = 18, units = "cm", dpi = 600)

```

## UMAP of neurons AFTER stage integration
See panel 1c for more information
```{r}
Stage.cols<- c("6"="#009E73", "24"="#CC79A7", "36"="#E69F00", "48"="#0072B2")

DimPlot(VNC.neurons, 
        cols = flybrainSeq::age.cols, 
        reduction = "umap", 
        group.by = "Stage", 
        pt.size = 1.5, 
        raster.dpi = c(1180,1180))&
  labs(title = "Neurons post-Integration")
ggsave(file.path(outputFolder, "neurons_post-integration.png"), width =20, height = 18, units = "cm", dpi = 600)
```

## Birthtime annotation of neurons
See panel 1d for more information
```{r}
birthorder_col <- c(earlysecondary="#4B458C", secondary="#B5C76A", primary="#4F642F", unknown="grey")
                  
DimPlot(VNC.neurons, 
        cols = birthorder_col, 
        reduction = "umap", 
        group.by = "birthtime", 
        pt.size = 1.5, 
        raster.dpi = c(1180,1180)) &
  labs(title = "Birth time annotation")
ggsave(file.path(outputFolder, paste0("birthtime_umap.png")), width = 20, height = 18, units = "cm", dpi = 600)
```

## Neurotransmitter annotation
See panel 2b for more information
```{r}
nt.cols <- c(GABA="cornflowerblue", Acetilcholine= "darkorange", Glutamate="#AB0D47", unknown="grey")

DimPlot(VNC.neurons, 
          reduction = "umap", 
          group.by = "neurotransmitter", 
          cols = nt.cols, 
          pt.size = 3, 
          raster=TRUE, 
          raster.dpi = c(4200,4200)) &
  labs(title = "Neurotransmitter annotation")
 ggsave(file.path(outputFolder, paste0("nt_identity_",red,".png")), 
         width = 20, height = 18, units = "cm", dpi = 1200)

```

## Hemilineage annotation
See panel 2c for more information
```{r}
EMhemilineagesColours <- c(
  "NA" = "#BEBEBE",
  "None" = "#BEBEBE",
  "TBD" = "#BEBEBE",
  "Unknown" = "#BEBEBE",
  "unassigned" = "#CCCCCC",
  "VGlut_orphan" = "black",
  "00A" = "#3B00FB",
  "00B" = "#00FA9A",
  "01A" = "#DEA0FD",
  "01B" = "#C075A6",
  "02A" = "#1C8356",
  "03A" = "#16FF32",
  "03B" = "#85660D",
  "04A" = "#D85FF7",
  "04B" = "#AA0DFE",
  "05A" = "#B5EFB5",
  "05B" = "#78B7C5",
  "06A" = "#3283FE",
  "06B" = "#B10DA1",
  "07B" = "#FA0087",
  "08A" = "#FF66FF",
  "08B" = "#BDCDFF",
  "09A" = "#C4451C",
  "09B" = "#822E1C",
  "10B" = "#2ED9FF",
  "11A" = "#E4E1E3",
  "11B" = "#90AD1C",
  "12A" = "#F7E1A0",
  "12B" = "#1CBE4F",
  "13A" = "#FEAF16",
  "13B" = "#325A9B",
  "14A" = "#FC1CBF",
  "14B" = "#D69C4E",
  "15A" = "#1E1E1E",
  "15B" = "#AAF400",
  "16B" = "#F6222E",
  "16X" = "#FB200E",
  "17A" = "#FBE426",
  "17B" = "#FAD510",
  "17X.18X" = "navy",
  "18B" = "#1C7F93",
  "19A" = "#F8A19F",
  "19B" = "#1CFFCE",
  "20A.22A" = "#B00068",
  "20A" = "#B00068",
  "22A" = "brown1",
  "20B.21B.22B" = "mediumaquamarine",
  "21A" = "#66B0FF",
  "21X" = "#5A5156",
  "23B" = "#9966CC",
  "24B.25B" = "#FAEFD1",
  "26X" = "#683B79",
  "27X" = "#8470FF"
)


DimPlot(VNC.neurons, reduction = "umap", group.by = "hemilineage",pt.size = 2, cols = EMhemilineagesColours, raster.dpi = c(1655,1655))&
  labs(title = "Hemilineage annotation")
ggsave(file.path(outputFolder, "VNC_combinedHemilineages.png"), width = 20, height = 18, units = "cm", dpi = 600)
```


## UMAP of primary and early secondary / secondary neurons
See panel 1c for more information
```{r}
#VNC.subset_allages <- readRDS(VNC.1ary.2ary.subset.folder)
VNC.subset_allages <- readRDS(VNC.1ary.2ary.subset.final.folder)

Stage.cols<- c("6"="#009E73", "24"="#CC79A7", "36"="#E69F00", "48"="#0072B2")

lapply(c("primary", "secondary"), function(Stage){
  DimPlot(VNC.subset_allages[[Stage]],
          cols = Stage.cols,
          reduction = "umap_subset",
          group.by = "Stage",
          pt.size = 1.5,
          raster.dpi = c(1180,1180)) &
    labs(title = paste(Stage, "neurons", sep=" "))
  ggsave(file.path(outputFolder, paste0(Stage,"_umap_subset.png")), width = 20, height = 18, units = "cm", dpi = 600)

})
```

